|

### 2024-03-25 index.qmd
update header
updaet
update manuscript

---

format:
  html:
    css: styles.css

---

@@

This ongoing project, guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/), aims to democratize access to longitudinal data analysis through practical examples and tutorials.

>>>>

---
---

@@

This ongoing project, guided by the Biostatistics Working Group of the [Adolescent Brain Cognitive Development (ABCD) Study](https://abcdstudy.org/), aims to provide open access to longitudinal data analysis through practical examples and tutorials.

### 2024-03-25 manuscript/manuscript.qmd
update manuscript

The size and scope of the ABCD Study data allow the research community to perform a large variety of developmental analyses of both substantive and methodological interest, presenting a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes unfold across critical periods of development. In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.0 data with accompanying R scripts. Worked examples are available in [Quarto](https://quarto.org/) files, accessible in the project's [GitHub repository]( https://github.com/OpenDevSci/LongDev-ABCD/tree/main/examples).

@@

The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged between 9-10 years at baseline, each with a parent/guardian. The study sample was recruited from household populations in defined catchment areas for each of the 21 (originally 22) study sites across the United States. Information regarding funding agencies, recruitment sites, investigators, and project organizations can be obtained at https://abcdstudy.org. The ABCD Study design is described in more detail in @garavan2018 and @dick2021.

@@

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior @moffitt2015. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified (e.g., @masten2005; @rogosch2010). Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold. Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors; e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants.

@@

Some longitudinal studies, e.g., accelerated longitudinal designs (ALDs; @thompson2011) are especially well suited for discovering these effects and modeling them. While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (9 to 11 years) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time). It is also possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, wherein number of prior assessments and age are mostly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

@@

The most common approach is to use random effects. Essentially, random effects allow for covariance estimates around fixed effects. A classic example (from @bryk1992; @singer1998) involves math achievement measured among students nested within schools. In a basic, intercept-only model with no covariates, there would be one fixed effect (the grand mean, or intercept, of math achievement), one school random effect (representing variation in the intercept between schools) and the within-school student residuals (variation left over after accounting for fixed and random effects). In this framework, each student’s score would be the sum of the fixed effect (the grand mean), the school random effect and the student’s within-school residual. Assumptions about the variance and covariance components of this model dictate the form of the variance/covariance structure. For example, if we assume the random effects are independent and identically distributed, the implied structure would be compound symmetry, where it is assumed the covariance of any two students in a single school is captured by a school random intercept and the covariance of any two students in different schools is zero. The assumptions of this relatively simple covariance structure can be relaxed depending on the nesting structure of the data, resulting in different covariance structures with additional parameters (see @singer1998).

@@

Modern approaches for handling missing data, such as full-information maximum likelihood, propensity weighting, auxiliary variables and multiple imputation avoid the biases of older approaches (see @enders2010; @graham2009). @graham2009 noted several “myths” regarding missing data. For example, Graham notes many assume the data must be minimally MAR to permit estimating procedures (such as maximum likelihood or multiple imputation) compared to other, more traditional approaches (e.g., using only complete case data). Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods. Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness (i.e., MCAR, MAR, MNAR; see @graham2009). The ABCD Biostatistics Workgroup is currently implementing several missing data approaches which are being implemented and compared to each other (and listwise deletion) in the 5.0 data release, including, propensity score weighting, and multiple (multilevel) imputation.

@@

An ideal longitudinal analysis integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model @collins2006. To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure 1). An understanding of these data structures is helpful, as they can warrant different types of LDA. Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and @bauer2019 provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data. Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change @bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest. This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events. While much has been written about “time-to-event” data @hosmer2008; @rizopoulos2012, including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study @brislin2023, our emphasis will be given to the modeling of “repeated measures” data.

@@

![LDA Table](img/Tabletemp.png){width=100% #fig-LDATable fig-alt="LDA Table" fig-align="left"}

>>>>

The size and scope of the ABCD Study data allow the research community to perform a large variety of developmental analyses of both substantive and methodological interest, presenting a unique opportunity to significantly advance our understanding of how a multitude of biopsychosocial processes unfold across critical periods of development. In this paper, we describe models and methods for longitudinal analysis of ABCD Study data that can address these fundamental scientific aims, including: 1) characterization of the genetic and environmental factors associated with variation in developmental trajectories; 2) assessment of how the level and timing of exposures may impact subsequent neurodevelopment; 3) quantification of how variation in developmental domains may be associated with outcomes, including mediation models and reciprocal relationships. We instantiate these longitudinal analyses in worked examples using the ABCD Release 5.1 data with accompanying R scripts. Worked examples are available in [Quarto](https://quarto.org/) files, accessible in the project's [GitHub repository](https://github.com/OpenDevSci/LongDev-ABCD).

@@

The ABCD Study enrolled a cohort of n=11,880 participants born between 2006-2008 and aged approximately 9-10 years at baseline, each with a parent/guardian. The study sample was recruited from households in defined catchment areas for each of the 21 (originally 22) study sites across the United States. Information regarding funding agencies, recruitment sites, investigators, and project organizations can be obtained at https://abcdstudy.org. The ABCD Study design is described in more detail in @garavan2018 and @dick2021.

@@

Normative development can also be upended by experiences (e.g., drug use) that, through various mechanisms, disrupt the normal flow of development wherein each stage establishes a platform for the next. For instance, substance use could lead to association with deviant peers, precluding opportunities for learning various adaptive skills and prosocial behaviors, in effect creating a “snare” that delays psychosocial development, such as maturing out of adolescent antisocial behavior @moffitt2015. Relatedly, the consequences of these types of events can cascade (e.g., school dropout, involvement in the criminal justice system) so that the effects of the snare are amplified (e.g., @masten2005; @rogosch2010). Although conceptually distinct from vulnerable periods, both types of developmental considerations highlight the importance of viewing behavior in the context of development and attempting to determine how various developmental pathways unfold. Longitudinal data are crucial in this context to assess individual levels of development prior to and following onset of experiences or other environmental factors (e.g., the ABCD Study collected data starting at ages 9-10 and hence before the onset of substance use for the vast majority of participants).

@@

Some longitudinal studies, e.g., accelerated longitudinal designs (ALDs; @thompson2011) are especially well suited for discovering these effects and modeling them. While ABCD is not an ALD, the variability in age (and grade in school) at the time of baseline recruitment (approximately 9-10 years old) allows some measures, collected every year, to be conceptualized as an ALD (e.g., substance use; prosocial behavior; family conflict; screen time). It is also possible that in later waves, analyses will allow for disaggregating the confounded effects of age and the number of prior assessments. However, ABCD is fundamentally a single-cohort, longitudinal design, wherein number of prior assessments and age are mostly confounded, and for, perhaps, most analyses, the possible influence of experience effects needs to be kept in mind.

@@

The most common approach is to use random effects. Essentially, random effects allow for covariance estimates around fixed effects. A classic example (from @bryk1992; @singer1998) involves math achievement measured among students nested within schools. In a basic, intercept-only model with no covariates (i.e., an unconditional growth model), there would be one fixed effect (the grand mean, or intercept, of math achievement), one school random effect (representing variation in the intercept between schools) and the within-school student residuals (variation left over after accounting for fixed and random effects). In this framework, each student’s score would be the sum of the fixed effect (the grand mean), the school random effect and the student’s within-school residual. Assumptions about the variance and covariance components of this model dictate the form of the variance/covariance structure. For example, if we assume the random effects are independent and identically distributed, the implied structure would be compound symmetry, where it is assumed the covariance of any two students in a single school is captured by a school random intercept and the covariance of any two students in different schools is zero. The assumptions of this relatively simple covariance structure can be relaxed depending on the nesting structure of the data, resulting in different covariance structures with additional parameters (see @singer1998).

@@

Modern approaches for handling missing data, such as full-information maximum likelihood, propensity weighting, auxiliary variables and multiple imputation avoid the biases of older approaches (see @enders2010; @graham2009). @graham2009 noted several “myths” regarding missing data. For example, Graham notes many assume the data must be minimally MAR to permit estimating procedures (such as maximum likelihood or multiple imputation) compared to other, more traditional approaches (e.g., using only complete case data). Violations of MAR impact both traditional and more modern data estimation procedures, though as noted by Graham, violations of MAR tend to have a greater effect on older methods. Graham thus suggests that imputing missing data is a better approach compared to listwise deletion in most circumstances, regardless of the model of missingness (i.e., MCAR, MAR, MNAR; see @graham2009; but also see @twisk2013). The ABCD Biostatistics Workgroup is currently implementing several missing data approaches which are being implemented and compared to each other (and listwise deletion) in the 5.0 data release, including, propensity score weighting, and multiple (multilevel) imputation.

@@

An ideal longitudinal analysis integrates (a) a well-articulated theoretical model, (b) an appropriate longitudinal data structure, and (c) a statistical model that is an operationalization of the theoretical model @collins2006. To accommodate various research questions and contexts, different types of longitudinal data and data structures have emerged (see Figure 1). An understanding of these data structures is helpful, as they can warrant different types of LDA. Given that identifying a starting point for making comparisons is somewhat arbitrary, Curran and @bauer2019 provide a nice on-ramp in first distinguishing between the use of “time-to-event” and “repeated measures” data. Although both model time, the former is concerned with whether and when an event occurs, whereas the later is focused on growth and change @bauer2019 Time-to-event structures measure time from a well-defined origin point up to the occurrence of an event of interest. This data structure is most often analyzed using survival analysis methods (e.g., hazard rate models, event history analysis, failure-time models and the time-to-event data can be based on a single assessment or include multiple recurrent or competing events). While much has been written about “time-to-event” data @hosmer2008; @rizopoulos2012, including a recent analysis examining exclusionary discipline in schools using data from the ABCD Study @brislin2023, our emphasis will be given to the modeling of “repeated measures” data.

@@
